Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_seminat.Rmd
revision of ndvi wetland, seminat & natopen bookdown-files

# NDVI- Semi-natural Ecosystems {#NDVI-indicator-seminat}

@@

```{r where-are-seminat-sites, fig.cap="A map of Norway showing the location of the semi-natural ecosystem sites."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31 - boreal heath",
      "T32 - semi-natural grasslands",
      "T33 - semi-natural coastal grasslands",
      "T34 - coastal heathlands",
      "T40/41 - previously heavily altered grassland"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

@@

```{r hist-sentinel-ndvis, fig.cap="A histogram showing the distribution of polygon sizes."}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

Semi-natural polygons stretch a large number of sizes from just above
100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
sqm.

@@

```{r NDVI-values-over-time-ndvis, fig.cap="A figure showing NDVI values for semi-natural ecosystem over time for six different main nature types in varying degrees of condition."}
SentinelNDVI.seminat %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

@@

```{r NDVI-major-basic-ndvis, warning=FALSE, fig.cap="NDVI values in major and basic semi natural ecosystem types in good condition."}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "C13", "C14", "C15", "C16", "C17", "C18", "C19", "C20", "C21", "E1", "E2", "E3", "E4", "E5", "E6", "E7"))
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
```

@@

**How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)**

```{r ndvi-regions-ndvis, warning=FALSE, fig.cap="DVI variation across regions in semi-natural ecosystems (major types) in good condition"}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1))+
  facet_wrap(~hovedtype)
```

@@

```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradien, faceted by six semi natural ecosystem types in five regions in Norway. "}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype)
```

@@

```{r north-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Northern Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Central Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r west-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in W-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in E-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r  south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in S-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets."}
# plot
allSatNDVI.seminat %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```

>>>>

# NDVI - Semi-natural Ecosystems {#NDVI-indicator-seminat}

NDVI (greenness) indicator, semi-natural ecosystems

@@

```{r where-are-seminat-sites, fig.cap="A map of Norway showing the location of the semi-natural ecosystem sites."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE, alpha=0) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31",
      "T32",
      "T33",
      "T34",
      "T40/41"
    ),
    title = "Major ecosystem type",
    size = 1.5
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems \n(NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```
T31...Boreal heath, T32...semi-natural grasslands, T33...semi-natural coastal grasslands, T34...Coastal heathlands, T40/T41...previously heavily managed grasslands (now similar to T32)

@@

```{r hist-sentinel-ndvis, fig.cap="A histogram showing the distribution of polygon sizes."}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
abline(v = 100, lty = 2)
```

Semi-natural polygons stretch a large number of sizes from just above
100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
sqm.
The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.

@@

```{r NDVI-values-over-time-ndvis, fig.cap="A figure showing NDVI values for semi-natural ecosystem over time for six different main nature types in varying degrees of condition."}
SentinelNDVI.seminat %>%
  filter(tilstand!="") %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Year")
```

@@

```{r NDVI-major-basic-ndvis, warning=FALSE, fig.cap="NDVI values in major and basic semi natural ecosystem types in good condition."}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  ggtitle("NDVI for different ecosystem types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")

SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "C13", "C14", "C15", "C16", "C17", "C18", "C19", "C20", "C21", "E1", "E2", "E3", "E4", "E5", "E6", "E7"))
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey" ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 7)) +
  ggtitle("NDVI for different ecosystem sub-types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
```

@@

**How does NDVI vary across regions in semi-natural ecosystems (major types) in good condition? (using only NDVI data matching NiN-mapping years)**

```{r ndvi-regions-ndvis, warning=FALSE, fig.cap="DVI variation across regions in semi-natural ecosystems (major types) in good condition"}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("NDVI for different ecosystems in good condition across regions") +
  labs(y = "NDVI values (Sentinel-2)", x = "Region")
```

@@

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
  ggtitle("NDVI for different ecosystems in different condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```

@@

```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradien, faceted by six semi natural ecosystem types in five regions in Norway. "}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(!is.na(region)) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```

@@

```{r north-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Northern Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Northern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Central Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Central Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

```{r west-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in W-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Western Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in E-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Eastern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

```{r  south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in S-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Southern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets."}
# plot
allSatNDVI.seminat %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  ggtitle("NDVI across time") +
  labs(y = "NDVI values", x = "Year") +
  facet_wrap(~hovedtype, ncol = 1)
```
